---
title: "GES660: Building Spatial Datasets"
knitr:
  opts_chunk: 
    echo: false
    message: false
editor_options: 
  markdown: 
    wrap: 72
---

Date 📅

:   August 31 -- December 7, 2022

Time ⏰

:   Wednesdays, 6:00 pm to 8:30 pm

Location 📍

:   Janet & Walter Sondheim Hall 001 (Cartography Lab), University of
    Maryland, Baltimore County, 1099 Hilltop Road, Baltimore, MD 21250

::: {.callout-note appearance="simple"}
This syllabus was updated on `r format(Sys.time(), '%B %d, %Y')`.
:::

## About this course

In this course, students will learn how to find, understand, and work
with spatial data for research and practice. This course leverages
open-source tools, online educational resources, and real-world data
from urban environments to help students build a methodological
framework for academic and professional work with spatial data now and
in the future.

The course explores the process of building and maintaining data sets
about local places and prepares students to navigate critical issues
including data ownership, the challenges of administrative data, and
privacy consideration. Assignments and readings will introduce students
to the range of uses for spatial data in planning, public policy, and
advocacy around housing, health, transportation, the environment, and
more.

Students in the course will learn to work with common file formats (e.g.
GeoJSON, GeoPackage) and web services (e.g. FeatureServer, APIs) and how
to read, write, document, and share data using QGIS, the R programming
language, and various web services.

::: {.callout-warning appearance="minimal"}
This course does not require any prior experience with desktop GIS
software or R programming. Assignments will require students to use R,
RStudio, and QGIS during the course. Optional resources on working with
spatial data using ArcGIS Online or Python will be provided where
feasible. Students will be required to use GitHub in order to share
completed assignments and develop their professional portfolios.
:::

## Schedule overview

Overall, this course is divided into three sections:

| **Section**                       | **Topics**                                                                                                                                  | **Sessions/Dates**          |
|----------------|----------------------------------------|----------------|
| Getting started with spatial data | Introduction to using R, sf, and QGIS to work with vector, raster, and tabular data.                                                        | 1 (Aug. 31) -- 6 (Oct. 5)   |
| Building and sharing spatial data | Downloading and manipulating spatial data from Socrata, ArcGIS, OpenStreetMap, and other sources using R.                                   | 7 (Oct. 12) -- 13 (Nov. 16) |
| Special topics and final project  | Use local data sources related to housing, transportation, community development, or related topics to work on a collaborative data project | 13 (Nov. 23) -- 15 (Dec. 7) |

Review [the course schedule](/schedule.html) for a more outline of
topics, readings, and activities.

## Assignments

The work for this course is split across three different activities:

-   Writing a weekly log in response to assigned readings

-   Completing lab assignments (primarily using R and QGIS)

-   Collaborating on a final project using spatial data

### Weekly log

Each week students are expected to write a 1-2 paragraph reflection the
assigned readings and 1-3 questions related to the technical materials.
Each weekly log entry should be committed to the "log" folder as a
Markdown (.md) file with a file name corresponding to the session date
(e.g. "2021-08-31_log.md"). Weekly logs that are unrelated to the
readings or do not include a question will be considered incomplete but
may be revised and resubmitted. You can miss up to four weekly log
entries during the course so each entry submitted beyond 10 entries is a
bonus point.

### Assignments

For each assignment, students will be provided instructions or reference
materials to complete a set of objectives. Assignments in R will be
submitted as an R script or RMarkdown file committed to the course
repository. Assignments in QGIS will typically be documented with a
brief written report and an exported map in a PDF format also submitted
using the course repository.

Assignments will include a clear list of objectives that must be met to
be considered complete. Most assignments may also include a bonus
objective that can earn additional points. Each assignment is
effecitvely graded along the following scale:

-   3: Assignment is submitted, complete, and includes a complete bonus
    assigmment

-   2: Assignment is submitted and complete

-   1: Assignment is submitted but incomplete

-   0: Assignment not submitted

Incomplete assignments can be revised and resubmitted for full credit.
Assignments submitted more than one day late cannot receive a bonus
point. Assignments more than one week late will not be accepted barring
exceptional circumstances.

```{r assignments}
knitr::kable(
tibble::tribble(
                                              ~Assignment, ~Session, ~Due.date,
                           "Complete GitHub Fundamentals",       2L,     "9/7",
                                      "Find spatial data",       2L,     "9/7",
             "Read spatial data and make a map with QGIS",       4L,    "9/21",
    "Read spatial data and make a map with R and ggplot2",       4L,    "9/21",
  "Compare spatial and geometry operations in QGIS and R",       5L,    "9/28",
                             "Document a public data set",       6L,    "10/5",
           "Tidy tabular address data and geocode with R",       7L,   "10/12",
                  "Georeference a historic map with QGIS",       8L,   "10/19",
   "Complete an exploratory analysis of a public dataset",       9L,   "10/26",
       "Make a map using OpenStreetMap data in QGIS or R",      13L,   "11/16"
  )
,
col.names = c("Assignment", "Session due", "Due date"),
align = "lcc"
)
```

### Final project

All students will participate in a final project that can be completed
independently or in collaboration with other students in the class. The
final project must use local spatial data and/or include the creation of
new spatial data about local places. Students will be encouraged to use
R and Quarto for the creation of the final project but projects
completed using QGIS are also acceptable.

A draft project proposal (1-2 pages in length) must be submitted by
November 9. We will review final projects as a group during the final
class session on December 7 and all project materials must be submitted
in a final format by December 14. Evaluation of the final project will
be primarily a self-assessment by the individual student or group with
the option for the instructor to add bonus points in recognition of
exceptional work or to adjust points awarded through self-assessment (in
rare cases).

## Evaluation

The evaluation of this course is intended to help you focus on
completing assignments---not getting them perfect.

```{r}
knitr::kable(
  tibble::tribble(
              ~Activity, ~Points, ~Bonus.available,
           "Weekly log",     10L,               4L,
          "Assignments",     20L,              10L,
     "Project proposal",      8L,               0,
        "Final project",     12L,               4,
     )
,
  col.names = c("Activity", "Points", "Bonus points"),
align = "rcc"
)

```

# **Statements & Policies**

## **Diversity Statement on Respect**

Students in this class are encouraged to speak up and participate during
class meetings. Because the class will represent a diversity of
individual beliefs, backgrounds, and experiences, every member of this
class must show respect for every other member of this class.

*From California State University, Chico's Office of Diversity and
Inclusion.*

## **Accessibility and Disability Accommodations, Guidance and Resources**

Accommodations for students with disabilities are provided for all
students with a qualified disability under the Americans with
Disabilities Act (ADA & ADAAA) and Section 504 of the Rehabilitation Act
who request and are eligible for accommodations. The Office of Student
Disability Services (SDS) is the UMBC department designated to
coordinate accommodations that creates equal access for students when
barriers to participation exist in University courses, programs, or
activities.

If you have a documented disability and need to request academic
accommodations in your courses, please refer to [the SDS
website](https://sds.umbc.edu/) at [sds.umbc.edu](http://sds.umbc.edu/)
for registration information and office procedures.

-   SDS email: disAbility\@umbc.edu
-   SDS phone: 410-455-2459

If you will be using SDS approved accommodations in this class, please
contact the instructor to discuss implementation of the accommodations.
During remote instruction requirements due to COVID, communication and
flexibility will be essential for success.

## **Sexual Assault, Sexual Harassment, and Gender Based Violence and Discrimination**

[UMBC
Policy](https://oei.umbc.edu/gender-discrimination-sexual-misconduct/)
and Federal law (Title IX) prohibit discrimination and harassment on the
basis of sex, sexual orientation, and gender identity in University
programs and activities. Any student who is impacted by sexual
harassment, sexual assault, domestic violence, dating violence,
stalking, sexual exploitation, gender discrimination, pregnancy
discrimination, gender-based harassment or retaliation should contact
the [University's Title IX
Coordinator](https://oei.umbc.edu/title-ix-coordinator/) to make a
report and/or access support and resources:

-   Morgan Thomas, Acting Director and Title IX Coordinator
-   410-455-1354 (direct line), morganthomas\@umbc.edu

You can access support and resources even if you do not want to take any
further action. You will not be forced to file a formal complaint or
police report. Please be aware that the University may take action on
its own if essential to protect the safety of the community.

If you are interested in or thinking about making a report, please use
the [Online Reporting/Referral
Form](https://umbc-advocate.symplicity.com/titleix_report/index.php/pid364290?).
Please note that, if you report anonymously,  the University's ability
to respond will be limited.

::: {.callout-important appearance="minimal"}
### **Faculty are Responsible Employees with Mandatory Reporting Obligations**

All faculty members are considered Responsible Employees, per [UMBC's
Policy on Sexual Misconduct, Sexual Harassment, and Gender
Discrimination.](https://oei.umbc.edu/policy-on-sexual-misconduct-sexual-harassment-and-gender-discrimination/)
Faculty are therefore required to report any/ all available information
regarding conduct falling under the Policy and violations of the Policy
to the Title IX Coordinator, even if a student discloses an experience
that occurred before attending UMBC and/or an incident that only
involves people not affiliated with UMBC.  Reports are required
regardless of the amount of detail provided and even in instances where
support has already been offered or received.

While faculty members want encourage you to share information related to
your life experiences through discussion and written work, students
should understand that faculty are required to report past and present
sexual assault, domestic and interpersonal violence, stalking, and
gender discrimination that is shared with them to the Title IX
Coordinator so that the University can inform students of their [rights,
resources and support](https://oei.umbc.edu/rights-and-resources/). 
While you are encouraged to do so, you are not obligated to respond to
outreach conducted as a result of a report to the Title IX Coordinator.

If you need to speak with someone in confidence, who does not have an
obligation to report to the Title IX Coordinator, UMBC has a number of
[Confidential
Resources](https://oei.umbc.edu/policy-on-sexual-misconduct-sexual-harassment-and-gender-discrimination/#confidential-resources)
available to support you:

-   [The Counseling Center](https://counseling.umbc.edu/) (Main Campus):
    410-455-2472 / After-Hours 410-455-3230 \[Monday -- Friday; 8:30
    a.m. -- 5 p.m.\]

-   [Center for Counseling and
    Consultation](https://shadygrove.umd.edu/student-services/center-for-counseling-and-consultation)
    (Shady Grove Campus): 301-738-6273 (Messages checked hourly) [Online
    Appointment Request Form](https://shadygrove.titaniumhwc.com/)

-   [University Health Services](https://uhs.umbc.edu/): 410-455-2542
    \[Monday -- Friday 8:30 a.m. -- 5 p.m.\]

-   Pastoral Counseling via [Interfaith
    Center](https://i3b.umbc.edu/spaces/the-interfaith-center-ifc/):
    410-455-3657; interfaith\@umbc.edu \[7 days a week; Fall and Spring
    7 a.m. -- 11 p.m.; Summer and Winter 8 a.m. -- 8 p.m.\]

Other Resources:

-   [Women's Center](https://womenscenter.umbc.edu/) (for students of
    all genders): 410-455-2714; womenscenter\@umbc.edu. \[Monday --
    Thursday 10:00am-5:30pm and Friday 10:00am-4pm\]

-   [Shady Grove Student
    Resources](https://oei.umbc.edu/shady-grove-title-ix-resources/),
    [Maryland Resources](https://oei.umbc.edu/maryland-resources/),
    [National Resources](https://oei.umbc.edu/national-resources/).
:::

::: {.callout-important appearance="minimal"}
### [**Child Abuse and Neglect**](https://oei.umbc.edu/child-protection/)

Please note that Maryland law and [UMBC
policy](https://oei.umbc.edu/child-protection/) require that faculty
report all disclosures or suspicions of child abuse or neglect to the
Department of Social Services and/or the police even if the person who
experienced the abuse or neglect is now over 18.
:::

## **Plagiarism**

Copying or using another's work in written or oral form---partial or
complete---without giving credit to the other person is a serious
academic offense and is taken very seriously in this class, by the
Department and by the University of Maryland, Baltimore County. UMBC
specifically defines plagiarism as anyone who "knowingly, or by
carelessness or negligence, representing as one's own in any academic
exercise the words, ideas, works of art or computer-generated
information and images of someone else."

Any student who plagiarizes will be referred to the Department Chair and
will be subject to the policies of the university. In general, the
consequences of plagiarism include failing an assignment, receiving a
lower course grade, and even failing a course. Examples of plagiarism
include:

-   Submit someone else's work as your own.
-   Buy a paper from a paper-mill, website or other source.
-   Copy sentences, phrases, paragraphs, or ideas from someone else's
    work, published or unpublished, without giving the original author
    credit.
-   Replace select words from a passage without giving the original
    author credit.
-   Copy any type of graphics, tables, graphs, maps, or charts from
    someone else's work without giving the original author credit.
-   Piece together phrases, ideas, and sentences from a variety of
    sources to write an essay.
-   Build on someone else's idea or phrase without giving the original
    author credit.

Details about avoiding [plagiarism, examples, and disciplinary
policies](http://aok.lib.umbc.edu/informationliteracy/plagiarism.php)
should be reviewed to gain a clear understanding prior to working on an
assignment or exam.

## **COVID-19 Safety Protocols and Compliance Statement**

UMBC has set clear expectations for masking while on campus that include
the requirement that you must wear a KN95 face mask or equivalent that
covers your nose and mouth in all classrooms regardless of your
vaccination status. Find more information on masks equivalent to KN95s
at <https://covid19.umbc.edu/masks/>

This is to protect your health and safety as well as the health and
safety of your classmates, instructor, and the university community.
Anyone attending class without a KN95 mask or wearing one improperly
will be asked by the instructor to put on a KN95 mask or fix their mask
in the appropriate position. Any student that refuses to comply with
this directive will be asked to leave the classroom immediately and
failure to do so may result in the instructor requesting the assistance
of the University Police. Students who refuse to wear KN95 masks may be
referred to [Student Conduct and Community
Standards](https://conduct.umbc.edu/) and may face disciplinary action
for violations of the Code of Student Conduct, specifically, Rule 2:
Behavior Which Jeopardizes the Health or Safety of Self or Others and
Rule 16: Failure to Comply with the Request of a University Official.
UMBC's on-campus safety protocols, including masking requirements, are
subject to change in response to the evolving situation with COVID-19.
